Estimating SARS-CoV-2 transmission in England from randomly sampled prevalence surveys

### 2023-12-20 transmission.tex
continue restructure

@@

The advantage of fitting a multilevel regression model on prevalence, rather than incidence, is that prevalence can naturally be expressed as a generalised linear model (GLM).
A natural formulation for either incidence proportion or prevalence in a stratum $i$ is $\logit(\pi_i) = g(t)$, where $\pi_i$ is the incidence proportion or prevalence in stratum $i$ and $g$ is a function of time, penalised towards being linear.
This is a GLM with a logit link function.
The logit transformation serves two purposes.
First, it ensures that the incidence proportion or prevalence is between 0 and 1.
Second, at low values of incidence or prevalence, linear growth is approximately exponential which is the natural assumption for an epidemic and has been shown to improve estimates of prevalence\todo{cite REACT paper on this, and check the exact meaning of "improve"}.
However, if the incidence proportion is expressed as a GLM in this way, the prevalence is not a GLM because a non-linear transformation is required to obtain prevalence from the logit of incidence proportion (the relationship on the natural scale is linear, as derived in \cref{E-inc-prev:sec:prevalence-process}).
Using a GLM is computationally useful due to the large number of pre-existing packages available for computing the posterior distribution, or its approximation, efficiently.
Previously, when fitting to prevalence over long periods and including all appropriate covariates to correct for non-response bias, the model was too computationally expensive to fit using MCMC; instead, an implementation of the model using the Integrated Nested Laplace Approximation (INLA) algorithm was used~\citePersonalComms{Koen Pouwels}.

@@

I used age groupings:\todo{insert age groupings}.
These align with previous work on the CIS\todo{cite CIS reports}.
They are based on school ages (primary, compulsory secondary, and post-compulsory secondary) for children to capture that school opening and closing may have a large effect on transmission.

@@

\section{Mechanistic modelling}

{\color{red} After introducing the observational model in generic terms above (section 8.2) I would see a reorganisation of the section 8.3 as follows:   
\begin{itemize}
    \item 8.3 (currently  titled Implementation) CIS-based transmission model (see comments on lines 484-486) 
\item 8.3.1 WAIFW matrix
\item 8.3.2 Full model (perhaps another title..model structure?) 
\item 8.3.3 Observational model (currently section 8.2) include here the "Solving the system of ODEs" (currently in section 8.3.4)
\item 8.3.4 Priors 
\item 8.3.5 Inference 
\item 8.4 Simulation study
\end{itemize}
}

\todo[inline]{Redo this intro, with some high-level comments taken from the full model section (eg following Birrell et al.)}

@@

A parsimonious model is needed to parameterise the WAIFW matrix (see \cref{SEIR:sec:structured-populations}).
I use the following age-based strata, which align closely with the age groupings previously used for analysis of the CIS data~\autocites[e.g.:][]{houseInferring}{walkerTracking}.
\todo{Contrast with age groupings used in previous chapter}
\begin{enumerate}
    \item $[0, 11)$: children of pre-school and primary school age.
    \item $[11, 16)$: children of compulsory secondary-school age.
    \item $[16, 25)$: adolescents and young adults who may remain in formal education.
    \item $[25, 50)$: younger working age adults.
    \item $[50, 70)$: older working age adults.
    \item $[70, \infty)$: elderly.
\end{enumerate}

I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, with the decomposition $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.
This separates contact patterns, $\kappa_{aa'}$, which can vary over time; transmission probabilities $q_{aa'}$, assumed constant; a region-wide effect for time $\beta(t)$; and the population of the stratum, $N_{a'}$.

@@

% Here, I extend the system of ODEs to include compartments representing PCR positive individuals.
% The proportion of PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are PCR positive as the mean of the Beta distribution.
One approach would be to convolve the incidence based on the transmission model with the PCR duration distribution.
Here incidence is calculated from the ODE solution and then related to the data as explained in \cref{E-inc-prev}.
However, this approach is computationally expensive because of the need to calculate the convolution in \cref{E-inc-prev:eq:EPt }.

A much more efficient option is to approximate this convolution by adding compartments to the ODEs.
This approach also allows the infected, pre-positive compartment, which I denote $p_0$, to be represented.
$p_0$ increases at the same rate as individuals flow into $e_1$.
These individuals then transition into a set of compartment with their transition rates chosen to match the duration of positivity estimates in \cref{E-imperf-test}, $p_1, \dots, p_l$.
These compartments do not have a biological interpretation, unlike the other compartments in the system, but are purely a mathematical tool so that the time spent in these compartments matches the duration distribution.
This is the same idea behind having multiple exposed or latent compartments (introduced in \cref{SEIR:sec:non-exponential}).

@@

The full model combines the elements discussed in this chapter up until now.
The transmission model follows \textcite{birrellRealtime}.
It is a SEIR model with two latent and two infectious states.
The population is stratified by age, using a time-varying matrix $\matr{\beta}(t)$ to describe the transmission between strata and how these change over time.

To form whole of England estimates, each region is modelled independently with the results aggregated, as in \cref{E-backcalc}.

The observation model is novel, relating the latent transmission described by the model to the CIS data.
It uses the phase-type model to represent the duration distribution of PCR positivity explained in \cref{SEIR:sec:observation}.

@@

\subsubsection{Solving the systems of ODEs} \label{SEIR:sec:solving-ODEs}
I solve the system of ODEs using Euler's method, discretised at timesteps of $\delta t = 1/2$ days.
That is, if $\vec{x}_t$ is the system state at time $t$ for $t \in \{ 0, 0.5, 1, \dots \}$ then $x_{t+\frac{1}{2}} = x_t + (d\vec{x}(t)/dt) \delta t$

@@

\subsubsection{Inference} \label{SEIR:sec:MCMC}

@@

Following \textcite{birrellBayesian}, I reparameterise the model to improve identifiability.
Rather than directly inferring $\beta_1$, I infer the initial growth rate of the epidemic.
In addition, I use a parsimonious parameterisation of the initial conditions using the initial proportion of the population that are infectious and the initial proportion that are susceptible only.
Both of these rely on an assumption that the epidemic is in a steady-state exponential growth phase at time 0 (see \cref{SEIR:sec:SIR}).

@@

Furthermore, $\R$ is a function of $\matr{\beta}$, and hence $\beta_1$ (as explained in \cref{SEIR:sec:structured-populations}).
Combining these expressions, the model can be parameterised by using $\psi$ in place of $\beta_1$.
The following procedure gives $\beta_1$ in terms of $\psi$.
First, calculate $\R$ from $\psi$ using \cref{SEIR:eq:rtoR}; denote this $\hat{R}$.
Then calculate $R^* = D(\matr{\beta}) / \gamma$ when $\beta_1=1$ (recall that $D(\matr{\beta})$ denotes the dominant eigenvalue of $\matr{\beta}$).
Multiplying a matrix by a scalar multiplies its dominant eigenvalue by that same scalar, therefore, setting $\beta_1 = \hat{R} / R^*$ gives the desired means that $\R$ at time 0 is $\hat{R}$, as desired.

@@

I ran a simulation study to check that the inference procedure (\cref{SEIR:sec:inference-implementation}) was working correctly and that the model was identifiable.
I ran 100 simulations, with parameters drawn from the priors (see \cref{SEIR:table:priors}) except that the prior on $\psi$ was $\LN(0.048, 0.0035)$.
I used the more informative prior for the simulation study to ensure that the simulations were realistic.
The weakly informative prior, used for the main analysis, allows a wide range of epidemics including some that are shrinking at the start of the time period.
Priors and population sizes were matched to the East of England region, over the same time period as the main analysis.

All fixed values, including $\matr{\kappa}$, used the previously described in this section.
This means that the simulations created realistic incidence and prevalence curves (see \cref{SEIR:fig:sim-data}).
In particular, the October half-term school holiday is visible, especially in the school-age groups and the November lockdown is visible to some extent in all age groups.
\begin{figure}
    \thisfloatpagestyle{empty}
    \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/sim/data}}
    \caption[Simulated data]{%
        Incidence (top) and prevalence (bottom) for the 100 simulated data sets as a proportion of each stratum.
        Prevalence shown is $\bar{p}_{kat}$, the true population prevalence on each day, excluding sampling noise.
        Each line represents one simulation.
    }
    \label{SEIR:fig:sim-data}
\end{figure}

@@

In this setup, the parameters $i^+$, $q_c$, and $\psi$ are fully recovered; the other parameters are only partially so: their posterior has moved away from the prior but is still regularised by it (see \cref{SEIR:fig:true-vs-posterior}).
The ability to identify $q_c$ is an advantage of using data which includes measurements on children, unlike data on hospitalisations or deaths.
This suggests that these parameters are not practically identifiable in this setting.
The posteriors align very closely with the priors (not shown) which reinforces this finding.
The lack of identifiability for $\vec{\pi}$ aligns with previous work showing this parameter is structurally unidentifiable, \ie not identifiable even in the limit of infinite data~\autocite{dankwaStructural}.
\begin{figure}
    \includegraphics{SEIR/sim/true_vs_posterior}
    \caption[True vs posterior parameter values]{%
        True vs posterior median parameter values for each parameter in each simulation.
        The black line shows the line $y = x$ which the dots would lie on if the parameter is fully recovered.
        The blue line is a linear regression line.
        For $i^+$, $q_c$, and $\psi$, the posterior recovers the true values.
        For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value.
    }
    \label{SEIR:fig:true-vs-posterior}
\end{figure}

@@

\section{Comparison of results}

>>>>

In this chapter, I use the estimate of the duration of PCR positivity from \cref{E-imperf-test} to estimate the transmission of SARS-CoV-2 in England.
This estimation is based on the CIS data (see \cref{transmission:sec:data} for further details).
Estimating transmission is the ultimate aim of this thesis.

I take two approaches.
The theoretical advantages of each approach were discussed in \cref{E-inc-prev:sec:infection-process}.
First, in \cref{backcalc}, I use a backcalculation approach, based on \cref{E-inc-prev:eq:EPt}.
Then, in \cref{SEIR}, I use a mechanistic compartmental model, extending the model introduced in \cref{E-SEIR:sec:mechanistic-models}.
Finally, in \cref{transmission:sec:comparison}, I compare the results of the two approaches.

\section{Data} \label{transmission:sec:data}

Throughout this chapter, the CIS data will be stratified into the regions of England and the following age groups.
\begin{enumerate}
    \item $[0, 11)$: children of pre-school and primary school age.
    \item $[11, 16)$: children of compulsory secondary-school age.
    \item $[16, 25)$: adolescents and young adults who may remain in formal education.
    \item $[25, 50)$: younger working age adults.
    \item $[50, 70)$: older working age adults.
    \item $[70, \infty)$: elderly.
\end{enumerate}
These groupings align closely with those previously used for analysis of the CIS data~\autocites[e.g.:][]{houseInferring}{walkerTracking}.
\todo{Contrast with age groupings used in previous chapter}

@@

The advantage of fitting a multilevel regression model on prevalence, rather than incidence, is that prevalence can naturally be expressed as a generalised linear model (GLM).
A natural formulation for either incidence proportion or prevalence in a stratum $i$ is $\logit(\pi_i) = g_i(t)$, where $\pi_i$ is the incidence proportion or prevalence in stratum $i$ and $g$ is a function of time, penalised towards being linear.
This is a GLM with a logit link function.
The logit transformation serves two purposes.
First, it ensures that the incidence proportion or prevalence is between 0 and 1.
Second, at low values of incidence or prevalence, linear growth is approximately exponential which is what mechanistic models predict, and has been shown to improve estimates of prevalence\todo{cite REACT paper on this, and check the exact meaning of "improve"}.
However, if the incidence proportion is expressed as a GLM in this way, the prevalence is not a GLM because a non-linear transformation is required to obtain prevalence from the logit of incidence proportion (the relationship on the natural scale is linear, as derived in \cref{E-inc-prev:sec:prevalence-process}).
Using a GLM is computationally useful due to the large number of pre-existing packages available for computing the posterior distribution, or its approximation, efficiently.
Previously, when fitting to prevalence over long periods and including all appropriate covariates to correct for non-response bias, the model was too computationally expensive to fit using MCMC; instead, an implementation of the model using the Integrated Nested Laplace Approximation (INLA) algorithm was used~\citePersonalComms{Koen Pouwels}.

@@

@@

\section{Mechanistic modelling} \label{SEIR}

As discussed in \cref{E-inc-prev:sec:infection-process}, there are various advantages to using a mechanistic model to estimate transmission.
This section uses a mechanistic model (introduced in \cref{E-SEIR:sec:mechanistic-models}) to estimate the transmission of SARS-CoV-2 in England.

The transmission part of the model follows \textcite{birrellRealtime}.
A novel approach (\cref{SEIR:sec:observation}) links the transmission to the data.
In the mechanistic modelling literature, this part of the model is referred to as the observation model, however, it includes both the prevalence process and the observation process (as defined in \cref{E-inc-prev}).

@@

A parsimonious model is needed to parameterise the WAIFW matrix (see \cref{E-SEIR:sec:structured-populations}).
I follow the WAIFW matrix parameterisation of \textcite{birrellRealtime}, with the decomposition $\beta_{aa'}(t) = \beta(t) \kappa_{aa'}(t) q_{aa'} N_{a'}$.
This separates contact patterns, $\kappa_{aa'}$, which can vary over time; transmission probabilities $q_{aa'}$, assumed constant; a region-wide effect for time $\beta(t)$; and the population of the stratum, $N_{a'}$.

@@

% Here, I extend the system of ODEs to include compartments representing PCR positive individuals.
% The proportion of PCR positive individuals in each strata is then linked to the CIS using a Beta-Binomial likelihood, with the modelled proportion that are PCR positive as the mean of the Beta distribution.
One approach would be to convolve the incidence based on the transmission model with the PCR duration distribution.
Here incidence is calculated from the ODE solution and then related to the data as explained in \cref{E-inc-prev}.
However, this approach is computationally expensive because of the need to calculate the convolution in \cref{E-inc-prev:eq:EPt}.

A much more efficient option is to approximate this convolution by adding compartments to the ODEs.
This approach also allows the infected, pre-positive compartment, which I denote $p_0$, to be represented.
$p_0$ increases at the same rate as individuals flow into $e_1$.
These individuals then transition into a set of compartment with their transition rates chosen to match the duration of positivity estimates in \cref{E-imperf-test}, $p_1, \dots, p_l$.
These compartments do not have a biological interpretation, unlike the other compartments in the system, but are purely a mathematical tool so that the time spent in these compartments matches the duration distribution.
This is the same idea behind having multiple exposed or latent compartments (introduced in \cref{E-SEIR:sec:non-exponential}).

@@

The remaining model structure follow \textcite{birrellRealtime}.
The transmission model is a SEIR model with two latent and two infectious states.
To form whole of England estimates, each region is modelled independently with the results aggregated, as in \cref{E-backcalc}.

@@

I solve the system of ODEs using Euler's method, discretised at timesteps of $\delta t = 1/2$ days.
That is, if $\vec{x}_t$ is the system state at time $t$ for $t \in \{ 0, 0.5, 1, \dots \}$ then $x_{t+\frac{1}{2}} = x_t + (d\vec{x}(t)/dt) \delta t$

@@

\subsubsection{Inference} \label{SEIR:sec:inference}

@@

Following \textcite{birrellBayesian}, I reparameterise the model to improve identifiability.
Rather than directly inferring $\beta_1$, I infer the initial growth rate of the epidemic.
In addition, I use a parsimonious parameterisation of the initial conditions using the initial proportion of the population that are infectious and the initial proportion that are susceptible only.
Both of these rely on an assumption that the epidemic is in a steady-state exponential growth phase at time 0 (see \cref{E-SEIR:sec:SIR}).

@@

Furthermore, $\R$ is a function of $\matr{\beta}$, and hence $\beta_1$ (as explained in \cref{E-SEIR:sec:structured-populations}).
Combining these expressions, the model can be parameterised by using $\psi$ in place of $\beta_1$.
The following procedure gives $\beta_1$ in terms of $\psi$.
First, calculate $\R$ from $\psi$ using \cref{SEIR:eq:rtoR}; denote this $\hat{R}$.
Then calculate $R^* = D(\matr{\beta}) / \gamma$ when $\beta_1=1$ (recall that $D(\matr{\beta})$ denotes the dominant eigenvalue of $\matr{\beta}$).
Multiplying a matrix by a scalar multiplies its dominant eigenvalue by that same scalar, therefore, setting $\beta_1 = \hat{R} / R^*$ gives the desired means that $\R$ at time 0 is $\hat{R}$, as desired.

@@

I ran a simulation study to check that the inference procedure (\cref{SEIR:sec:inference}) was working correctly and that the model was identifiable.
I ran 100 simulations, with parameters drawn from the priors (see \cref{SEIR:table:priors}) except that the prior on $\psi$ was $\LN(0.048, 0.0035)$.
I used the more informative prior for the simulation study to ensure that the simulations were realistic.
The weakly informative prior, used for the main analysis, allows a wide range of epidemics including some that are shrinking at the start of the time period.
Priors and population sizes were matched to the East of England region, over the same time period as the main analysis.

All other aspects of the simulation model, including $\matr{\kappa}$, are as previously described in this section.
This means that the simulations created realistic incidence and prevalence curves (see \cref{SEIR:fig:sim-data}).
In particular, the October half-term school holiday is visible, especially in the school-age groups and the November lockdown is visible to some extent in all age groups.
\begin{figure}
    \thisfloatpagestyle{empty}
    \vspace{-3cm}
    \makebox[\textwidth][c]{\includegraphics[width=.9\paperwidth]{SEIR/sim/data}}
    \caption[Simulated data]{%
        Incidence (top) and prevalence (bottom) for the 100 simulated data sets as a proportion of each stratum.
        Prevalence shown is $\bar{p}_{kat}$, the true population prevalence on each day, excluding sampling noise.
        Each line represents one simulation.
    }
    \label{SEIR:fig:sim-data}
\end{figure}

@@

\todo[inline]{What happened to the following paragraph? It doesn't make sense currently!}
In this setup, the parameters $i^+$, $q_c$, and $\psi$ are fully recovered; the other parameters are only partially so: their posterior has moved away from the prior but is still regularised by it (see \cref{SEIR:fig:true-vs-posterior}).
The ability to identify $q_c$ is an advantage of using data which includes measurements on children, unlike data on hospitalisations or deaths.
This suggests that these parameters are not practically identifiable in this setting.
The posteriors align very closely with the priors (not shown) which reinforces this finding.
The lack of identifiability for $\vec{\pi}$ aligns with previous work showing this parameter is structurally unidentifiable, \ie not identifiable even in the limit of infinite data~\autocite{dankwaStructural}.
\begin{figure}
    \includegraphics{SEIR/sim/true_vs_posterior}
    \caption[True vs posterior parameter values]{%
        True vs posterior median parameter values for each parameter in each simulation.
        The black line shows the line $y = x$ which the dots would lie on if the parameter is fully recovered.
        The blue line is a linear regression line.
        For $i^+$, $q_c$, and $\psi$, the posterior recovers the true values.
        For $\vec{\pi}$ and $\rho$, the posterior median is very similar regardless of the true value.
    }
    \label{SEIR:fig:true-vs-posterior}
\end{figure}

@@

\section{Comparison of results} \label{transmission:sec:comparison}
